### 2017-07-07 motivation.tex
Sec. 2 edits.

In this paper, we adopt a different approach that \emph{lifts}
isolation semantics (\emph{not} their implementations) to the
application layer, providing a principled framework to simultaneously
reason about application invariants and isolation properties.  To
illustrate this idea informally, consider how we might verify that
\C{new\_order} is sound when executed under \emph{Snapshot Isolation}
(SI), and isolation level stronger than RC. Snapshot isolation
allows transactions to be executed against their respective private
snapshots of the database, thus admitting concurrency, but it also
requires that there not be any write-writeconflicts among
concurrent transactions. Write-write conflicts can be eliminated in various
ways; either through detection followed by a rollback, or through
exclusive locks, or a combination of both, leading to different ways
in which SI can be realized. For instance, one possible implementation
of SI, close to the one use by PostgreSQL~\cite{postgres-ssi},
executes a transaction against its private snapshot of the database,
but obtains exclusive locks on the actual records in the database
before performing writes. A write is performed only of the record
hasn't already been updated by a concurrent transaction, i.e,  when it
doesn't result in a write-write conflict.  Otherwise, the transaction is
rolledback. This implementation is demonstrated for a simple database
with three records, $a$, $b$, and $c$ in Fig.~\ref{fig:RR-postgres}.

@@

Implementations of SI on real databases (e.g., PostgreSQL) are
complicated, often running into thousands of lines of code.
Nonetheless, the semantics of Snapshot Isolation, in terms of how it
effects the transitions on the database state, can be captured in a
fairly simple model. Firstly, although SI admits concurrent
transactions, their effects are  not witnessed in the current
transaction (call it $T$), during $T$'s execution. Thus, insofar as
$T$ is concerned, as long as it executes, the global state doesn't
change. More formally, for every consecutive pair of global states
$(\stg,\stg')$ witnessed by $T$ during its execution, $\stg'=\stg$.
After $T$ finishes execution, it commits to the actual database, which
incorporates the effects of concurrent transactions. In executions
where $T$ successfully commits, concurrent transactions are guaranteed
to not be in write-write conflict with $T$. Thus, if $\stg$ is the
global state that $T$ witnessed when it finished execution (which
would be same as the snapshot state), and $\stg'$ is the state to
which $T$ commits, then the diff between $\stg$ and $\stg'$ shouldn't
be in a write-write conflict with $T$. To concretize this notion, let
the database state be a map from transaction variables to values, and
let $\stl$ denote a transaction-local log that maps the variables
being written to their updated values. Then the absence of write-write
conflicts between $T$ and the diff between $\stg$ and $\stg'$ can be
defined as: $\forall x\in\mathit{dom}(\delta)$, $\stg'(x) = \stg(x)$.
To summarize, the semantics of SI can be captured as an axiomatization
over transitions of the database state ($\Delta \longrightarrow
\Delta'$) during the lifetime an SI transaction ($T$):
\begin{itemize}
  \item While $T$ executes, $\Delta' = \Delta$.
  \item After $T$ finishes execution, but before it commits its local
    state $\delta$, $\forall(x\in\delta).~\Delta'(x) = \Delta(x)$.
\end{itemize}

@@

This simple characterization of SI isolation allows us to verify the
consistency conditions associated with the \C{new\_order} transaction.
First, since the database doesn't change ($\Delta' = \Delta$) during
the execution of the transaction's body, we can reason about
\C{new\_order} as though it executed in complete isolation until its
commit point, leading to a verification process similar to what would
been applied when reasoning about serializability.  When
\C{new\_order} finishes execution, but before it commits, the SI
axiomatization shown above requires us to consider global state
transitions $\stg \longrightarrow \stg'$ that do not include changes
to the records ($\stl$) written by \C{new\_order}, i.e.,
$\forall(x\in\delta).~\Delta'(x) = \Delta(x)$. Since concurrent
\C{new\_order} transactions that write to the same \C{District} record
modify the record by incrementing its \C{d\_next\_o\_id} field, their
transitions do not satisfy the necessary condition imposed by the SI
axiomatization, allowing us to rightfully ignore their interference.
This leaves the interference due to concurrent \C{new\_order}
transactions that write to a different district record.  Fortunately,
such interference does not prompt \C{new\_order} to violate the
invariant, which can be established by adopting a reasoning similar to
the one used on concurrent program logics, such as the Rely-Guarantee
logic, customized for the database programs.

@@

Fig.~\ref{fig:foreach_code} shows a (simplified) snippet of code taken
from Fig.~\ref{fig:new_order_code}. Some irrelevant expressions have
been replaced with constants (\C{k1} to \C{k3}).  The body of the loop
executes a SQL update followed by an insert.  Recall that a transaction
reads from the global database ($\Delta$), and writes to a
transaction-local database ($\delta$) before committing these updates. An update
statement filters the tuples that match the search criteria from $\Delta$
and computes the updated tuples that are to be
added to the local database. Thus, its state transformer (call it
$T_U$) is the following function on sets\footnote{$\bind$ has higher precedence than $\cup$.}:
\begin{ocaml}
  fun ($\Delta$,$\delta$) -> $\delta$ $\cup$ $\Delta$$\bind$(fun s -> if table(s) = stock && s.s_i_id = item_req.ol_i_id 
                                 then {{s with s_qty = k1}} (* a singleton set *)
                                 else $\emptyset$)
\end{ocaml}
% \begin{smathpar}
% \begin{array}{c}
%   \lambda\Delta.\lambda\delta.\, \delta \;\cup\; 
%       \Delta \,\bind\, (\lambda s.\, 
%           \C{if}\;{\C{stock}(s) \conj 
%                    \C{s.s\_i\_id} = \C{item\_req.ol\_i\_id}}\\
%            \hspace*{0.9in}
%            \C{then}\;{\{\{s \;\C{with}\; \C{s\_qty} = \C{k1}\}\}}\;
%            \C{else}\;{\emptyset})
% \end{array}
% \end{smathpar}
% \begin{verbatim}
%   Rmem(σ'(s')) = Rmem(σ(s')) U 
%                  Rmem[\x. Rmem[\y. if y<x then {y+x} else {}](σ(s)) U {x}](xs)
% \end{verbatim}
Observe that
$T_U(\Delta,\delta)$ is of the form $\delta \,\cup\,
F_U(\Delta)$. The transformer ($T_I(\Delta,\delta)$)
for the subsequent \C{insert} statement can be similarly calculated to
be of the form $\delta \cup F_I(\Delta)$.
Their composition gives the state transformer of the loop body:
\begin{ocaml}
  fun ($\Delta$,$\delta$) -> $\delta$ $\cup$ $F_U(\Delta) \,\cup\, F_I(\Delta)$
\end{ocaml}
Let $F_{body}(\Delta) = F_U(\Delta) \,\cup\, F_I(\Delta)$.   The
transformer for the \C{foreach} loop can now be computed as follows:
\begin{ocaml}
  fun ($\Delta$,$\delta$) -> $\delta$ $\cup$ item_reqs$\bind$(fun item_req -> $F_{body}(\Delta)$)
\end{ocaml}
Observe that the transformer captures the precise semantics of the
loop as a formula in the domain of sets extended with a bind
function. The advantage in doing so is that we can now make use of a
semantics-preserving translation from this domain to first-order
logic, allowing us to leverage SMT solvers for automatic proofs
without having to infer loop invariants, or provide intermediate
assertions.  Sec.~\ref{sec:automation} describes this translation. In
the exposition thus far, we assumed $\Delta$ remains invaraint, which
is clearly not the case when we admit concurrency.  Necessary
concurrency extensions of the state transformer semantics is also
covered in Sec.~\ref{sec:automation}.  The following two sections,
however, first focus on laying the theoretical foundations for
reasoning about weak isolation.

>>>>

In this paper, we adopt a different approach that \emph{lifts}
isolation semantics (\emph{not} their implementations) to the
application layer, providing a principled framework to simultaneously
reason about application invariants and isolation properties.  To
illustrate this idea informally, consider how we might verify that
\C{new\_order} is sound when executed under \emph{Snapshot Isolation}
(SI), and isolation level stronger than RC. Snapshot isolation
allows transactions to be executed against their respective private
snapshots of the database, thus admitting concurrency, but it also
requires that there not be any write-writeconflicts among
concurrent transactions. Write-write conflicts can be eliminated in various
ways; either through detection followed by a rollback, or through
exclusive locks, or a combination of both, leading to different ways
in which SI can be realized. For instance, one possible implementation
of SI, close to the one use by PostgreSQL~\cite{postgres-ssi},
executes a transaction against its private snapshot of the database,
but obtains exclusive locks on the actual records in the database
before performing writes. A write is performed only if the record
hasn't already been updated by a concurrent transaction, i.e,  when it
doesn't result in a write-write conflict.  Otherwise, the transaction is
rolledback. This implementation is demonstrated for a simple database
with three records, $a$, $b$, and $c$ in Fig.~\ref{fig:RR-postgres}.

@@

Implementations of SI on real databases (e.g., PostgreSQL) are
complicated, often running into thousands of lines of code.
Nonetheless, the semantics of Snapshot Isolation, in terms of how it
effects the transitions on the database state, can be captured in a
fairly simple model. Firstly, although SI admits concurrent
transactions, their effects are  not witnessed in the current
transaction (call it $T$), during $T$'s execution. Thus, insofar as
$T$ is concerned, the global state doesn't change as long as it
executes. More formally, for every consecutive pair of global states
$(\stg,\stg')$ witnessed by $T$ during its execution, SI requires
$\stg'=\stg$.  After $T$ finishes execution, it commits to the actual
database, which incorporates the effects of concurrent transactions.
In executions where $T$ successfully commits, concurrent transactions
are guaranteed to not be in write-write conflict with $T$. Thus, if
$\stg$ is the global state that $T$ witnessed when it finished
execution (which would be same as the snapshot state), and $\stg'$ is
the state to which $T$ commits, then the diff between $\stg$ and
$\stg'$ shouldn't be in a write-write conflict with $T$. To concretize
this notion, let the database state be a map from transaction
variables to values, and let $\stl$ denote a transaction-local log
that maps the variables being written to their updated values. Then
the absence of write-write conflicts between $T$ and the diff between
$\stg$ and $\stg'$ can be defined as: $\forall
x\in\mathit{dom}(\delta)$, $\stg'(x) = \stg(x)$.  To summarize, the
semantics of SI can be captured as an axiomatization over transitions
of the database state ($\Delta \longrightarrow \Delta'$) during the
lifetime an SI transaction ($T$):
\begin{itemize}
  \item While $T$ executes, $\Delta' = \Delta$.
  \item After $T$ finishes execution, but before it commits its local
    state $\delta$, $\forall(x\in\delta).~\Delta'(x) = \Delta(x)$.
\end{itemize}

@@

This simple characterization of SI isolation allows us to verify the
consistency conditions associated with the \C{new\_order} transaction.
First, since the database doesn't change ($\Delta' = \Delta$) during
the execution of the transaction's body, we can reason about
\C{new\_order} as though it executed in complete isolation until its
commit point, leading to a verification process similar to what would
been applied when reasoning about serializability.  When
\C{new\_order} finishes execution, but before it commits, the SI
axiomatization shown above requires us to consider global state
transitions $\stg \longrightarrow \stg'$ that do not include changes
to the records ($\stl$) written by \C{new\_order}, i.e.,
$\forall(x\in\delta).~\Delta'(x) = \Delta(x)$. Since concurrent
\C{new\_order} transactions that write to the same \C{District} record
modify the record by incrementing its \C{d\_next\_o\_id} field, their
transitions do not satisfy the necessary condition imposed by the SI
axiomatization, allowing us to rightfully ignore their interference.
This leaves the interference due to concurrent \C{new\_order}
transactions that write to a different district record.  Fortunately,
such interference does not prompt \C{new\_order} to violate the
invariant, which can be established by adopting a reasoning similar to
the one used on concurrent program logics, such as the Rely-Guarantee
logic, customized for the database programs. Our approach generalizes
this style of reasoning to a range of isolation levels, concurrency
control techniques, and their combinations used frequently in practice.

@@

Fig.~\ref{fig:foreach_code} shows a (simplified) snippet of code taken
from Fig.~\ref{fig:new_order_code}. Some irrelevant expressions have
been replaced with constants (\C{k1} to \C{k3}).  The body of the loop
executes a SQL update followed by an insert.  Recall that a transaction
reads from the global database ($\Delta$), and writes to a
transaction-local database ($\delta$) before committing these updates. An update
statement filters the tuples that match the search criteria from $\Delta$
and computes the updated tuples that are to be
added to the local database. Thus, its state transformer (call it
$\T_U$) is the following function on sets\footnote{$\bind$ has higher
precedence than $\cup$.}:
\begin{smathpar}
\begin{array}{l}
  \lambda(\stl,\stg).~ \stl \cup \stg \bind(\lambda s. 
      \itel{{\sf table}(s) = \C{Stock} \conj 
                s.\C{s\_i\_id} = \C{item\_req.ol\_i\_id}\\\hspace*{1.3in}} 
           {\{\{s \with \C{s\_qty} = \C{k1}\}\}\\\hspace*{1.3in}}
           {\emptyset}
\end{array}
\end{smathpar}
% \begin{smathpar}
% \begin{array}{c}
%   \lambda\Delta.\lambda\delta.\, \delta \;\cup\; 
%       \Delta \,\bind\, (\lambda s.\, 
%           \C{if}\;{\C{stock}(s) \conj 
%                    \C{s.s\_i\_id} = \C{item\_req.ol\_i\_id}}\\
%            \hspace*{0.9in}
%            \C{then}\;{\{\{s \;\C{with}\; \C{s\_qty} = \C{k1}\}\}}\;
%            \C{else}\;{\emptyset})
% \end{array}
% \end{smathpar}
% \begin{verbatim}
%   Rmem(σ'(s')) = Rmem(σ(s')) U 
%                  Rmem[\x. Rmem[\y. if y<x then {y+x} else {}](σ(s)) U {x}](xs)
% \end{verbatim}
% Observe that
% $\T_U(\Delta,\delta)$ is of the form $\delta \,\cup\,
% F_U(\Delta)$. 
The transformer ($\T_I(\Delta,\delta)$) for the subsequent \C{insert}
statement can be similarly calculated as the following:
\begin{smathpar}
\begin{array}{l}
  \lambda(\stl,\stg).~ \stl \cup \{\C{ol\_o\_id}=\C{k2};\, 
        \C{ol\_d\_id}=\C{k3};\, \C{ol\_i\_id}=\C{item\_req.ol\_i\_id};\, 
        \C{ol\_qty}=\C{item\_req.ol\_qty}\}
\end{array}
\end{smathpar}
Observe that both transformers are of the form $T(\Delta,\delta) =
\stl \cup F(\stg)$, where $F$ is a function that returns the set of
records added by the transformer to the transaction-local database
($\stl$). Let $\F_U$ and $\F_I$ be the corresponding functions for $\T_U$
and $\T_I$ shown above. The state transformed of the loop body in
Fig.~\ref{fig:new} can be expressed as the following composition of
$\F_U$ and $\F_I$:
\begin{smathpar}
\begin{array}{l}
  \lambda(\stl,\stg).~ \stl \cup \F_U(\stg) \cup \F_I(\stg)
\end{array}
\end{smathpar}
The transformer for the loop itself can now be computed as following:
\begin{smathpar}
\begin{array}{l}
  \lambda(\stl,\stg).~ \stl \cup \C{item\_reqs}\bind
      (\lambda\C{item\_req}.~ \F_U(\stg) \cup \F_I(\stg))
\end{array}
\end{smathpar}
Observe that the structure of the transformer mirrors the structure of
the program itself. In particular, SQL statements become set
operations, and the \C{DB} monad's $\bind$ becoms set monadic $\bind$.
Thus, state transformer inference is completely mechanical. The
advantage of inferring such transformers is that we can now make use
of a semantics-preserving translation from the domain of sets equipped
with $\bind$ to the first-order logic, allowing us to leverage SMT
solvers for automatic proofs without having to infer loop invariants,
or provide intermediate assertions.  Sec.~\ref{sec:inference}
describes this translation. In the exposition thus far, we assumed
$\Delta$ remains invaraint, which is clearly not the case when we
admit concurrency.  Necessary concurrency extensions of the state
transformer semantics is also covered in Sec.~\ref{sec:inference}.
The following two sections, however, first focus on laying the
theoretical foundations for reasoning about weak isolation.
